White Wine Quality Analysis by Suliman Rashed
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(knitr)
library(scales)
library(dplyr)
library(gridExtra)
library(GGally)
library(RColorBrewer)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
df <- read.csv('wineQualityWhites.csv')
```

This dataset is public and available for research purposes only. </br>
The link to the website is here: http://www3.dsi.uminho.pt/pcortez/wine/.
<br/>Citation:
P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. Modeling wine preferences by data mining from physicochemical properties. In Decision Support Systems, Elsevier, 47(4):547-553, 2009.

```{r echo=FALSE}
str(df)
```
 > The dataset has 13 variables and it's divided into two parts. First, the
 physicochemical section with 11 variables. Second, the quality variable for the
 score of the expert from 0 to 10. 
 </br> To use the quality variable as classification, a factor version of this
 variable is needed.
 
```{r}
df$quality.factor <- as.factor(df$quality)
str(df)
```

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
summary(df)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots2}
#to creat a function about histogram to help make grid plots faster
grid.plot <- function(variable, dataframe, label.x)
{
  ggplot(aes_string(variable), data = dataframe) + 
    geom_histogram(bins = 40) +
    geom_vline(aes(xintercept = mean(variable)), color = 'blue') +
    geom_vline(aes(xintercept = median(variable)), color = 'red') +
    geom_vline(aes(xintercept = quantile(variable, 0.01)),
               linetype = 2, color = 'orange') +
    geom_vline(aes(xintercept = quantile(variable, 0.99)),
               linetype = 2, color = 'orange') +
    geom_vline(aes(xintercept = quantile(variable, 0.25)), linetype = 2) +
    geom_vline(aes(xintercept = quantile(variable, 0.75)), linetype = 2) +
    labs(x = label.x)
}
# cal the function to creat plots
grid.p1 <- grid.plot(df$fixed.acidity, df, 'fixed.acidity')
grid.p2 <- grid.plot(df$volatile.acidity, df, 'volatile.acidity')
grid.p3 <- grid.plot(df$citric.acid, df, 'citric.acid')
grid.p4 <- grid.plot(df$pH, df, 'pH')
grid.p5 <- grid.plot(df$residual.sugar, df, 'residual.sugar')
grid.p6 <- grid.plot(df$chlorides, df, 'chlorides')
grid.p7 <- grid.plot(df$free.sulfur.dioxide, df, 'free.sulfur.dioxide')
grid.p8 <- grid.plot(df$total.sulfur.dioxide, df, 'total.sulfur.dioxide')
grid.p9 <- grid.plot(df$density, df, 'density')
grid.p10 <- grid.plot(df$sulphates, df, 'sulphates')
grid.p11 <- grid.plot(df$alcohol, df, 'alcohol')
grid.p12 <- grid.plot(df$quality, df, 'quality')

grid.arrange(grid.p1,grid.p2, grid.p3, grid.p4, grid.p5, grid.p6, grid.p7,
             grid.p8, grid.p9, grid.p10, grid.p11, grid.p12, ncol = 4)

```

> To clarify these charts: </br>
- Blue Line: Specify the median. </br>
- Red Line: Specify the mean. </br>
- Dashed Black Line: Specify the quantiles. </br>
- Dashed Orange Line: Specify the percentile (0.01 & 0.99). </br>
  We can see that all of these measures have a (almost) normal distribution and
  some outliers but doesn't affect the distribution.

 Since we have a categorical variable for `quality`, let's focus on divide it
into 3 categorical specifications (A is the best & C is the worst).

```{r echo=TRUE, message=FALSE, warning=FALSE, Univariate_Plots3}
# creating a new variable (quality catagories) "A" between 10-8, "B" between 7-5
# "C" between 4-0
df$quality.catagories <- ifelse(df$quality >= 8, "A",
                                ifelse(df$quality >= 5, "B",
                                       ifelse(df$quality >= 0, "C",
                                                            "Other")))
```

 After that, we will use one of these categorical specifications to draw plots
with physicochemical variables to know the specifications of the best white wine.

```{r echo=TRUE, message=FALSE, warning=FALSE, Univariate_Plots31}
ggplot(aes(df$quality.catagories), data = df) +
  geom_histogram(stat = "count") 
```

Above is a histogram that plot the `qualit.catagories`.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots4}
# creating subset data with quality catagory = A
dfbest <- subset(df, quality.catagories == "A")
#to creat a function about density plot to help make grid plots faster
best.plots <- function(variable, label.x)
{
  ggplot(aes_string(variable), data = dfbest) + 
    geom_density(fill = 'gray') +
    geom_vline(aes(xintercept = mean(variable)), color = 'blue') +
    geom_vline(aes(xintercept = median(variable)), color = 'red') +
    geom_vline(aes(xintercept = quantile(variable, 0.01)), 
               linetype = 2, color = 'orange') +
    geom_vline(aes(xintercept = quantile(variable, 0.99)),
               linetype = 2, color = 'orange') +
    geom_vline(aes(xintercept = quantile(variable, 0.25)), linetype = 2) +
    geom_vline(aes(xintercept = quantile(variable, 0.75)), linetype = 2) +
    labs(x = label.x)
}

best.p1 <- best.plots(dfbest$fixed.acidity, 'fixed.acidity')
best.p2 <- best.plots(dfbest$volatile.acidity, 'volatile.acidity')
best.p3 <- best.plots(dfbest$citric.acid, 'citric.acid')
best.p4 <- best.plots(dfbest$pH, 'pH')
best.p5 <- best.plots(dfbest$residual.sugar, 'residual.sugar')
best.p6 <- best.plots(dfbest$chlorides, 'chlorides')
best.p7 <- best.plots(dfbest$free.sulfur.dioxide, 'free.sulfur.dioxide')
best.p8 <- best.plots(dfbest$total.sulfur.dioxide, 'total.sulfur.dioxide')
best.p9 <- best.plots(dfbest$density, 'density')
best.p10 <- best.plots(dfbest$sulphates, 'sulphates')
best.p11 <- best.plots(dfbest$alcohol, 'alcohol')

grid.arrange(best.p1,best.p2, best.p3, best.p4, best.p5, best.p6, best.p7,
             best.p8, best.p9, best.p10, best.p11, ncol = 4)

```

It appears that most of the features for the best wine showed a normal
distribution except the `residual.sugar` and `alcohol`. Also there are some
positive outliers in almost each of these specifications.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots5}
#to creat a function about density plot to help make grid plots faster
sugar.plots <- function(variable, dataframe, label.x)
{
  ggplot(aes_string(variable), data = dataframe) + 
    geom_density(fill = 'gray') +
    geom_vline(aes(xintercept = mean(variable)), color = 'blue') +
    geom_vline(aes(xintercept = median(variable)), color = 'red') +
    geom_vline(aes(xintercept = quantile(variable, 0.01)),
               linetype = 2, color = 'orange') +
    geom_vline(aes(xintercept = quantile(variable, 0.99)),
               linetype = 2, color = 'orange') +
    geom_vline(aes(xintercept = quantile(variable, 0.25)), linetype = 2) +
    geom_vline(aes(xintercept = quantile(variable, 0.75)), linetype = 2) +
    labs(x = label.x)
}

sugar.p1 <- sugar.plots(df$residual.sugar, df, 'residual.sugar')
sugar.p2 <- sugar.plots(df$residual.sugar, df, 'residual.sugar log10') +
  scale_x_log10()

grid.arrange(sugar.p1, sugar.p2)
```

 As we can see there is a long tail in the first plot and most of the data are
under 20. When we did Log10 on the second plots you can spot the difference
clearly.


# Univariate Analysis


### What is the structure of your dataset?
The dataset contains 13 variables (11 physicochemical variables, one categorical
variable and one variable for indexing) and 4898 observations.
The 11 physicochemical features are: `fixed.acidity`, `volatile.acidity`, 
`citric.acid`, `residual.sugar`, `chlorides`, `free.sulfur.dioxide`, 
`total.sulfur.dioxide`, `density` ,`pH`, `sulphates`, and `alcohol`. and all of
them are numeric, while the remaining variable is `quality` and even when it is
written with numbers but it represented as categorical since it is ratings by
the experts.

### What is/are the main feature(s) of interest in your dataset?
The main feature is the `quality` variable since it represent the ratings of the
experts.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
There is `chlorides`, `density`, and `alcohol`.

### Did you create any new variables from existing variables in the dataset?
Yes. `quality.catagorical` from `quality`.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
In `residual.sugar` there was a long tail with most of the data were under 20.
So I did a log 10 transformation on its x-axis to see the transformation and it
was very clear.

# Bivariate Plots Section


```{r echo=FALSE, message=FALSE, Bivariate_Plots}

# create subset for the data by removing X variable and quality (factor & catagories)
tmpdata <- select(df, -X, -quality.factor, -quality.catagories)

ggcorr(tmpdata, label = TRUE, label_size = 3, size = 4, layout.exp = 5,
       hjust = 0.9, label_round = 2)

```


 We can see from the plots matrix above that the strongest correlation is 
between `quality` and `alcohol` with (r=0.436). Also, we can spot that the
correlation between `density` and `residual.sugar` is strong positive with
(r=0.839) and `density` and `total.sulfur.dioxide` with (r=0.53). And in terms
of negative correlation, the correlation between `density` and `alcohol` is 
(r=-78), and between `density` and `quality` is (r=-0.307).

```{r echo=FALSE, message=FALSE,warning=FALSE, Bivariate_Plots2}
#to creat a function about box plot to help make grid plots faster
boxPlots <- function(yvariable, dataframe, labs.y)
{
  ggplot(aes_string(x = df$quality.factor, y = yvariable, fill = df$quality.factor), data = dataframe) + 
    geom_boxplot() +
    ylim(quantile(yvariable, 0), quantile(yvariable, 0.99)) +
    theme(legend.position = "none") +
      labs(x = "quality.factor", y = labs.y)
}
# caling the function by each features to plot them
box.p1 <- boxPlots(df$fixed.acidity, df, "fixed.acidity")
box.p2 <- boxPlots(df$volatile.acidity, df, "volatile.acidity")
box.p3 <- boxPlots(df$citric.acid, df, "citric.acid")
box.p4 <- boxPlots(df$pH, df, "pH")
box.p5 <- boxPlots(df$residual.sugar, df, "residual.sugar")
box.p6 <- boxPlots(df$chlorides, df, "chlorides")
box.p7 <- boxPlots(df$free.sulfur.dioxide, df, "free.sulfur.dioxide")
box.p8 <- boxPlots(df$total.sulfur.dioxide, df, "total.sulfur.dioxide")
box.p9 <- boxPlots(df$density, df, "density")
box.p10 <- boxPlots(df$sulphates, df, "sulphates")
box.p11 <- boxPlots(df$alcohol, df, "alcohol")

grid.arrange(box.p1, box.p2, box.p3, box.p4, box.p5, box.p6, box.p7, box.p8, 
             box.p9, box.p10, box.p11, ncol = 3)
```

 As we can see from the matrix box plots that there are two features that really
affect the quality and the ratings of the wine. `alcohol` with a strong positive
and `density` with a strong negative. Other features don't have that much effect
as these two with `quality`. Let's focus on it.

```{r echo=FALSE, message=FALSE,warning=FALSE, Bivariate_Plots3}
grid.arrange(box.p9, box.p11)
```

From the `density` plot we can see that if the density is high we expect to 
see the rating lower. On the other side, as long as the `alcohol` is high we
expect the rating of the quality is high.

```{r echo=FALSE, message=FALSE,warning=FALSE, Bivariate_Plots4}
ggplot(aes(df$density, df$alcohol), data = df) +
  geom_point(color = '#374B5E', alpha = 1/5, position = "jitter") +
  ylim(quantile(df$alcohol, 0.1), quantile(df$alcohol, 0.99)) +
  xlim(quantile(df$density, 0.1), quantile(df$density, 0.99)) +
  geom_smooth(method = 'lm', color = '#061F37')
```

```{r echo=FALSE, message=FALSE,warning=FALSE, Bivariate_Plots5}
cat("r =", cor(df$density, df$alcohol))
```


Lets plot some of the other chimical features to see the relationship between them:

##### density and residual.sugar

```{r echo=FALSE, message=FALSE,warning=FALSE, Bivariate_Plots6}
ggplot(aes(df$density, df$residual.sugar), data = df) +
  geom_point(color = '#374B5E', alpha = 1/5, position = "jitter") +
  ylim(quantile(df$residual.sugar, 0.1), quantile(df$residual.sugar, 0.99)) +
  xlim(quantile(df$density, 0.1), quantile(df$density, 0.99)) +
  geom_smooth(method = 'lm', color = '#061F37')
```

```{r echo=FALSE, message=FALSE,warning=FALSE, Bivariate_Plots7}
cat("r =", cor(df$density, df$residual.sugar))
```

##### density and total.sulfur.dioxide

```{r echo=FALSE, message=FALSE,warning=FALSE, Bivariate_Plots8}
ggplot(aes(df$density, df$total.sulfur.dioxide), data = df) +
  geom_point(color = '#374B5E', alpha = 1/5, position = "jitter") +
  ylim(quantile(df$total.sulfur.dioxide, 0.1),
       quantile(df$total.sulfur.dioxide, 0.99)) +
  xlim(quantile(df$density, 0.1), quantile(df$density, 0.99)) +
  geom_smooth(method = 'lm', color = '#061F37')
```

```{r echo=FALSE, message=FALSE,warning=FALSE, Bivariate_Plots9}
cat("r =", cor(df$density, df$total.sulfur.dioxide))
```

##### alcohol and residual.sugar


```{r echo=FALSE, message=FALSE,warning=FALSE, Bivariate_Plots10}
ggplot(aes(df$alcohol, df$residual.sugar), data = df) +
  geom_point(color = '#374B5E', alpha = 1/5, position = "jitter") +
  ylim(quantile(df$residual.sugar, 0.1), quantile(df$residual.sugar, 0.99)) +
  geom_smooth(method = 'lm', color = '#061F37')
```

```{r echo=FALSE, message=FALSE,warning=FALSE, Bivariate_Plots11}
cat("r =", cor(df$alcohol, df$residual.sugar))
```

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
> We saw above that there is a strong correlation between `quality` and `alcohol`
with (r=0.436). Also, we saw that the correlation between `density` and
`residual.sugar` is strong positive with(r=0.839) and `density` and
`total.sulfur.dioxide` with (r=0.53). And in terms of negative correlation,
the correlation between `density` and `alcohol` is (r=-78), and between `density`
and `quality` is (r=-0.307).

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
> Yes. there was between `alcohol` and `density` and between `residual.sugar` and
`density` also.

### What was the strongest relationship you found?
> It was between `residual.sugar` and `density` with (r = 0.839).

# Multivariate Plots Section


```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots}
ggplot(aes(df$density, df$alcohol, color = df$quality.catagories), data = df) +
  geom_point(stat = "summary", fun.y = "median", alpha = 1) +
  ylim(quantile(df$alcohol, 0.1), quantile(df$alcohol, 0.99)) +
  xlim(quantile(df$density, 0.1), quantile(df$density, 0.99)) +
  geom_smooth(method = 'lm', color = '#061F37') +
  labs(title = "Median of Alcohol for Density By Quality")
```

With a strong negative relationship between `alcohol` and `density`, we can see
from the plot above that the best wine tend to have much alcohol, and the worst
tend to have higher density.

```{r echo=FALSE,message=FALSE, warning=FALSE, Multivariate_Plots2}
dprc <- ggplot(aes((2*(df$density/2)), df$residual.sugar, color = df$quality.catagories), data = df) +
  geom_jitter(stat = "summary", fun.y = "median") +
  ylim(min(df$residual.sugar), quantile(df$residual.sugar, 0.99)) +
  xlim(min(df$density), quantile(df$density, 0.99)) 

dpra <- ggplot(aes((2*(df$density/2)), df$residual.sugar, color = df$alcohol), data = df) +
  geom_jitter(stat = "summary", fun.y = "median") +
  ylim(min(df$residual.sugar), quantile(df$residual.sugar, 0.99)) +
  xlim(min(df$density), quantile(df$density, 0.99))

grid.arrange(dprc, dpra, ncol = 2)
```

With holding `density` in both plots, we can see in the left plot that a better
wine have more residual sugar. On the right side we can see also that an 
alcoholic wine have more residual sugar.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
> We can see that the best wine tend to have much alcohol, and the worst tend to
have higher density. ALso, we can see that a better wine have more residual 
sugar and an alcoholic wine have more residual sugar.

### Were there any interesting or surprising interactions between features?
> No, I didn't see anything.

------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
box.p9 +
  labs(title = "Density By Quality") +
  xlab("Quality") +
  ylab("Density (g / cm^3)")
```

### Description One
Since the correlation between `density` and `quality` is negative with 
(r = -0.3071233) we can see from the plot above that when the density is high we
expect to see the quality to be low.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(aes(df$density, df$residual.sugar), data = df) +
  geom_point(color = '#374B5E', alpha = 1/5, position = "jitter") +
  ylim(quantile(df$residual.sugar, 0.1), quantile(df$residual.sugar, 0.99)) +
  xlim(quantile(df$density, 0.1), quantile(df$density, 0.99)) +
  geom_smooth(method = 'lm', color = '#061F37') +
  labs(title = "Residual Sugar By Density") +
  xlab("Density (g / cm^3)") +
  ylab("Residual Sugar (g / dm^3)")
```

### Description Two

We can spot from the plot that the correlation between `residual.sugar` and 
`density` is string positive with (r = 0.838). So as long the residual sugar is
heigh we excpect the density to be higher.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE,Plot_Three}
ggplot(aes(df$density, df$alcohol, color = df$quality.catagories), data = df) +
  geom_point(stat = "summary", fun.y = "median", alpha = 1) +
  ylim(quantile(df$alcohol, 0.1), quantile(df$alcohol, 0.99)) +
  xlim(quantile(df$density, 0.1), quantile(df$density, 0.99)) +
  geom_smooth(method = 'lm', color = '#061F37') +
  labs(title = "Median of Alcohol for Density By Quality") +
  xlab("Density (g/ml)") +
  ylab("Alcohol (%)") +
  guides(color=guide_legend(title="Quality"))
```

### Description Three

With a strong negative relationship between `alcohol` and `density` 
(r = -0.7801376), we can see from the plot above that the best wine tend to have
much alcohol, and the worst tend to have higher density.

------

# Reflection

It was an interesting project and an interesting dataset to investigate it.
The gradation of the project was fascinating, starts from univariate to bivariate
to multivariate. Knowing which chemical property has the highest effect on wine's
quality. Also between the properties itself and how much that have an effect on 
the quality in the end. </br>
What I think it would give more accurate is to have more observations \ wines to
test since we don't have any wines that rated with 10 \ 10. </br>
In the future, we could implement these modules on other sets, maybe the red
wines that in the description of this set. Also, we may go deeper into this data 
set and explore more relation between the chemical properties.